---
title: 'N741: Exploratory Data Analysis'
author: "Melinda K. Higgins, PhD."
date: "January 30, 2017"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# initial R chunk - set global knitr options
# for now echo=TRUE so ALL R chunks will be shown
# in your final output.
#
# When you are ready to do a final document NOT showing
# the R code - run this with echo = FALSE instead.
knitr::opts_chunk$set(echo = TRUE)

# In this inital R chunk, this is the best place to go ahead
# and load ALL of the packages your R markdown document
# will need. Then you can then not have to worry about adding 
# the library() commands later on.
# Tis also helps anyone reading your code to know
# right at the beginning which packages are needed to
# execute your code and compile your document. Thus, they
# can go ahead and install the packages needed (which is done
# outside of and BEFORE the RMD file is compiled)

# packages needed for this RMD file
library(tidyverse)

```

## Worked Example from the UCI Data Repository

The following dataset comes from the [UCI Data Repository](http://archive.ics.uci.edu/ml/). The dataset we'll use is the Contraceptive Method Choice dataset. The information on this dataset is provided at [http://archive.ics.uci.edu/ml/datasets/Contraceptive+Method+Choice](http://archive.ics.uci.edu/ml/datasets/Contraceptive+Method+Choice). If you click on the "Data Folder" you can download the RAW data `cmc.data` which is a comma delimited format dataset (i.e. it is a CSV formatted file) and the description of the data included, the variable names and associated codes for the values included which is in the `cmc.names` file. See "Data Folder"" at [http://archive.ics.uci.edu/ml/machine-learning-databases/cmc/](http://archive.ics.uci.edu/ml/machine-learning-databases/cmc/)

### Read-in data

**NOTE:** Download the 2 files from the UCI Data Repository for the Contraceptive Method Choice and put them in the directory where you have this RMD `rmarkdown` file.

```{r}
# read in the comma delimited (CSV) formatted dataset
# **NOTE**: This dataset does NOT have the column
# names as the 1st row of the file. We will assign the 
# column names below.
cmc <- read_csv("cmc.data", col_names=FALSE)
```

### Apply the codebook - variable names and coding used

Apply variable names to the 10 columns of data in `cmc`.

```{r}
# assign new variables names to the 10 columns
names(cmc) <- c("WifeAge", "WifeEd", "HusbEd", "NumChild",
                "WifeRel", "WifeWork", "HusbOcc", "SOLindex", 
                "Media", "Contraceptive")
```

The next code chunk is to add the labels for "factor" levels for some of the variables (i.e. we are creating factors).

**WARNING**: Notice I'm overwriting the variables and changing them from integers to factors which have different properties as you'll see below. If you want to keep the original integer variables, you could simply give the new facotr variable a new name. For example you could write 

```
cmc$WifeEd.f <- factor(cmc$WifeEd,
                       levels = c(1,2,3,4),
                       labels = c("low","med low","med
```

and this would append a new column onto the `cmc` dataset that is the "factor" type version of Wife's Education. For now, use the code below to update all of the variables.

```{r}
# update Wife Education as a factor, assign the 
# levels and the labels for each level
cmc$WifeEd <- factor(cmc$WifeEd,
                     levels = c(1,2,3,4),
                     labels = c("low","med low","med high","high"))

# do the remaining variables
cmc$HusbEd <- factor(cmc$HusbEd,
                     levels = c(1,2,3,4),
                     labels = c("low","med low","med high","high"))

cmc$WifeRel <- factor(cmc$WifeRel,
                      levels = c(0,1),
                      labels = c("Non-Islam","Islam"))

# Note: The documentation does state that
# 0=yes and 1=no which seems incorrect...
cmc$WifeWork <- factor(cmc$WifeWork,
                       levels = c(0,1),
                       labels = c("Yes","No"))

cmc$HusbOcc <- factor(cmc$HusbOcc,
                      levels = c(1,2,3,4),
                      labels = c("1","2","3","4"))

cmc$SOLindex <- factor(cmc$SOLindex,
                       levels = c(1,2,3,4),
                       labels = c("low","med low","med high","high"))

cmc$Media <- factor(cmc$Media,
                    levels = c(0,1),
                    labels = c("Good","Not Good"))

cmc$Contraceptive <- factor(cmc$Contraceptive,
                            levels = c(1,2,3),
                            labels = c("No-use","Long-term","Short-term"))
```

### Look at a subset of the data

```{r}
head(cmc)
```

### Print this subset using `knitr::kable()`

```{r}
knitr::kable(head(cmc))
```

### Summarize the dataset

**NOTICE** that Wife's Age and Number of Children are now the only "numeric" "integer" variables - these are the only ones for which we get summary statistics. All the remaining variables are "factors" so we only get the frequencies for each category.

```{r}
summary(cmc)
```

### Computing stats on factors

Suppose you wanted to know the mean education level of the Huband's in this dataset. We can use the `as.numeric()` function to convert the variable and then run a `mean()` on it. We'll do more on facotrs later this semester.

```{r}
mean(as.numeric(cmc$HusbEd))
```

### Cleaning up your tables & Improving Workflow with PIPES (`%>%`)

```{r}
# these lines of code use the %>% "pipe" command.
# It also uses the group_by() function
# also in the dplyr package. The lines below can be read as
# "take the XXX (cmc) dataset, THEN summarise the
# sample size and sample mean.

# initial steps
cmc %>%
    summarise(nChild = length(NumChild),
              meanChild = mean(NumChild))

# look at output - figure out number of columns, add knitr::kable()
# put in good column names (2 columns) and a TITLE using caption
cmc %>%
    summarise(nChild = length(NumChild),
              meanChild = mean(NumChild)) %>%
  knitr::kable(col.names=c("N","mean"),
                 caption="Number of Children: Descriptive Stats")

# let's add more descriptive stats to our table
# this means we now have more columns - one per stat
cmc %>%
    summarise(nChild = length(NumChild),
              minChild = min(NumChild),
              meanChild = mean(NumChild),
              sdChild = sd(NumChild),
              medianChild = median(NumChild),
              maxChild = max(NumChild)) %>%
    knitr::kable(col.names=c("N","min",
                             "mean","sd","median","max"),
                 caption="Number of Children: Descriptive Stats")

# let's do again but BY Wife's Religion (2 groups)
# group the data BY
# each continent THEN summarise each continent's mean and sd."
# I THEN sent the output to the kable function to output
# there is one more column now for Wife's Religion.
cmc %>%
    group_by(WifeRel) %>%
    summarise(nChild = length(NumChild),
              minChild = min(NumChild),
              meanChild = mean(NumChild),
              sdChild = sd(NumChild),
              medianChild = median(NumChild),
              maxChild = max(NumChild)) %>%
    knitr::kable(col.names=c("Wife Religion","N","min",
                             "mean","sd","median","max"),
                 digits = 2,
                 caption="Number of Children: Stats by Wife Religion")

```

### Using and Getting R Code from `Rcmdr` package (with GUI interface)

OUTSIDE of the RMD file - run `library(Rcmdr)`. We'll load the `cmc` dataset into the local environment. Then use the GUI in `Rcmdr` to generate the R code for what you might want.

For example, getting the counts and relative %s for Wife Working categories.

```{r}
local({
  .Table <- with(cmc, table(WifeWork))
  cat("\ncounts:\n")
  print(.Table)
  cat("\npercentages:\n")
  print(round(100*.Table/sum(.Table), 2))
})
```

```{r}
Table <- with(cmc, table(WifeWork))
knitr::kable(as.data.frame(Table))
```

# two-way table % chi-square test

Open `Rcmdr` and run a two-way table with a Chi-square test of independence between Husband Occupation and Contraceptive use. And run the Rmarkdown - Generate Report from within Rcmdr to see how that report is set-up.

```{r}
# r packages Rcmdr Rmarkdown uses - you'll need these too
library(Rcmdr)
library(car)
library(RcmdrMisc)
library(rgl)
```

### capture R code chunks for the two-way table and chi-square stats

```{r}
library(abind, pos=23)
```

```{r}

# the following is the code generated by Rcmdr
local({
  .Table <- xtabs(~HusbOcc+Contraceptive, data=cmc)
  cat("\nFrequency table:\n")
  print(.Table)
  cat("\nRow percentages:\n")
  print(rowPercents(.Table))
  .Test <- chisq.test(.Table, correct=FALSE)
  print(.Test)
})
```

Let's pull out the code pieces we need and use `knitr` to make a better table.

```{r}
# create a Table object for the results from xtabs()
Table <- xtabs(~HusbOcc+Contraceptive, data=cmc)

# use the rowPercents (from the RcmdrMisc package)
# to pull these out of the Table
# make it a data.frame and make a nice table with knitr
knitr::kable(as.data.frame(rowPercents(Table)),
             caption = "Contraceptive Use by Husband Occupation")
```

### Chi-square test results

```{r}
# and run the chi-square test and show the results
chisq.test(Table, correct=FALSE)
```

### Make Plots with `Rcmdr`

Clustered Bar plot - side-by-side - Contraceptive Use by Husband Occupation

```{r}
with(cmc, Barplot(Contraceptive, by=HusbOcc, style="parallel", 
  legend.pos="topright", xlab="Contraceptive", ylab="Frequency"))
```

### Make a Clustered Bar plot with `ggplot2`

using the "dodge" option

```{r}
ggplot(cmc, aes(x=Contraceptive, fill=HusbOcc)) + 
  geom_bar(position='dodge')
```

using the "stack" option

```{r}
ggplot(cmc, aes(x=Contraceptive, fill=HusbOcc)) + 
  geom_bar(position='stack')
```

using the "fill" option

```{r}
ggplot(cmc, aes(x=Contraceptive, fill=HusbOcc)) + 
  geom_bar(position='fill')

```

### Scatterplot of Wife's Age and Number of Children

Remember there are 1473 subjects in this dataset.

```{r}
cmc %>%
  ggplot(aes(x=WifeAge, y=NumChild)) +
    geom_point()
```

This obviously has a lot of overplotting (points on top of one another). One way to alleviate this issue is to add "jitter" or a little bit of randomness so the points won't lie on top of one another.

```{r}
cmc %>%
  ggplot(aes(x=WifeAge, y=NumChild)) +
    geom_point(position = "jitter")
```

Still another way to "bin" the data in 2 dimensions with a lot of points in a scatterplot is to use the `geom_hex()` function which basically does a density plot using 2-D bins like a 2-D histogram in a way.

```{r}
cmc %>%
  ggplot(aes(x=WifeAge, y=NumChild)) +
    geom_hex()
```

### Boxplot options

```{r}
# make a boxplot showing the range of Wife Ages by
# Number of Children - "bin" widths = 1 child
# we'll use cut_width
cmc %>%
  ggplot(aes(x=NumChild, y=WifeAge)) +
    geom_boxplot(aes(group=cut_width(NumChild, 1)))

# we could also use a binning option based
# on n groups of equal Number of Children range
# let's try 8 intervals using cut_interval
cmc %>%
  ggplot(aes(x=NumChild, y=WifeAge)) +
    geom_boxplot(aes(group=cut_interval(NumChild, 8)))

# this time use cut_number and 6 bins
# so that each bin has about the same # of cases
cmc %>%
  ggplot(aes(x=NumChild, y=WifeAge)) +
    geom_boxplot(aes(group=cut_number(NumChild, 6)))
```


### Histograms and Density estimates of Continuous Data

Let's look at Wife's Age and Number of Children - do you expect these to look normal?

see more at [http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/](http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/)

```{r}
cmc %>%
  ggplot(aes(WifeAge)) +
    geom_histogram(aes(y=..density..),
                   colour="black",fill="white",
                   binwidth=5) +
    geom_density()

cmc %>%
  ggplot(aes(WifeAge)) +
    geom_histogram(aes(y=..density..),
                   colour="black",fill="yellow",
                   binwidth=5) +
    geom_density(alpha=.2, fill="blue")



```

What about overlaying a Normal Curve? Also add some better axis labels and a title

```{r}
cmc %>%
  ggplot(aes(WifeAge)) +
    geom_histogram(aes(y=..density..),
                   colour="black",fill="yellow",
                   binwidth=5) + 
    stat_function(fun = dnorm, 
                  args = list(mean = mean(cmc$WifeAge), 
                              sd = sd(cmc$WifeAge)), 
                  lwd = 1, 
                  col = 'red') +
  labs(title = "Distribution of Wife's Age",
       x = "Wife's Age",
       y = "Density")
```


